The evolution of gene expression in the term placenta of viviparous mammals

### 2016-02-18 manuscript/manuscript.Rnw
add changes suggested by derek

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental
mammals”. The term “placental mammals” is misleading, as extant
yinothereians (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, in which the maternal and fetal blood supply is separated by
one tissue layer, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
\DLA{To discuss with derek}

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Nanospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

                             "/Gene/Summary?g=",name_or_id,"}{",
                             xtable.sanitize.text(name_or_id),"}")]

@@

         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Symbol`),"}")]

@@

    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","FPKM SEM"))
    if (temp[,all(is.na(`FPKM SEM`))]) {
        temp[,`FPKM SEM`:=NULL]

@@

placenta.lineage.specific.table <- function(analysis,output=TRUE) {
    temp.table <- 
        head(placenta.lineage.specific.analysis[analysis.name==analysis &
                                                p.value <= 0.05][order(p.value),
                                                                 list(human_name,p.value,fdr,fc),],
             n=5)
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        print(xtable(temp.table[order(`p (t-test)`),],
                     display=c("s","s","s","e","g","f"),
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

    }
    return(invisible(temp.table[order(`p (t-test)`),]))
}
output.top.genes.table <- function(data,output=TRUE) {
    if (output) {
        table.display <- c("s","s","s","d")
        table.align <- c("c","c","p{3in}","r")
        if (any(colnames(data) %in% "FPKM SEM"))  {
            table.display <- c(table.display,"d")
            table.align <- c(table.align,"r")
        }
        if (any(colnames(data) %in% "Human Symbol"))  {
            table.align <- c("c",table.align)
            table.display <- c("s",table.display)
        }
        print(xtable(data,
                     display=table.display,
                     align=table.align
                     ),
              booktabs=TRUE,
              floating=FALSE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)
    }
}

@@

\begin{figure}
<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=16,fig.height=16,warning=FALSE,message=FALSE>>=
plot.label.size <- 3

pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2)))
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
                 fontsize=plot.label.size
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
                 fontsize=plot.label.size
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 vp=viewport(layout.pos.col=1,layout.pos.row=2),
                 fontsize=plot.label.size
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSSSCG00000006792",
                           "ENSBTAG00000046085"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=2),
                 fontsize=plot.label.size
                 )

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (GH1 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
GH2
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
the hormone adrenomedullin (\textit{ADM}).

@@

For the term placenta of \smonkey, assembled contigs from RNA-seq were
aligned using against \human and the best alignment was chosen as the
annotation for each assembled contig. In the term placenta of
\smonkey, the top 50 non-housekeeping genes
(\cref{tab:top50_spider_monkey_placenta}) include a possible regulator
of androgen receptor signaling
\cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
(\Sexpr{gene.expression("ateles_fusciceps","PAGE4")}), %
the pregnancy specific glycoprotein \textit{PSG8}
(\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
a member of the chorionic somatotropins, \textit{GH1} %
(\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
\cref{tab:top50_spider_monkey_placenta}), and multiple components of
annexin complexes
(\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("ateles_fusciceps","S100P",complete=TRUE)}, %
\Sexpr{gene.expression("ateles_fusciceps","S100A11",complete=TRUE)}) %
which may play a role in angiogenesis, and certainly regulate
cell-cell interaction in the developing placenta
\cite{Liu.ea2015:AnnexinA2complexeswith}.
\Sexpr{gene.expression("ateles_fusciceps","AL365202.1",complete=TRUE)}
is also highly expressed, has no known function, and is duplicated
multiple times in \textit{Macaca mulatta}. All genes with expression
greater than 50 \ac{FPKM} are listed in Supplementary
\cref{tab:all_ateles_fusciceps_placenta}.

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \carmeli are shown in
\cref{tab:top50_s_carmeli_placenta} with a list of all highly
expressed genes in Supplemental
\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
ware highly expressed
(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Hbb-a1",complete=TRUE)}, and
\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
well as the relaxin
\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
multiple cathespins
(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), and many prolactin-related genes
(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
expansion seen in \mouse.

@@

We examined the expression of five Laurasiatherians: \cow, \sheep,
\pig, \horse, and \dog. We were unable to detect any significant
differences after correcting for multiple testing between
Laurasiatherians and all other analyzed species
(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
Laurasiatherians and Euasiatherians
(Supplemental \cref{tab:laurasiatheria_eurasiatherian_lineage_specific}).

@@

\subsection{Atlantogenata}
\subsubsection{\elephant Placenta Transcriptome}

@@

\begin{figure}
<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
             aes(y=human_name,x=capfirst(Species)))
      + geom_raster(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low = grad.def[["low"]],
                            high = grad.def[["high"]],
                            space = grad.def[["space"]],
                            na.value = grad.def[["na.value"]],
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + ylab("Human Ortholog Name")
      + xlab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(), 
              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
      )
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes expressed (\ac{FPKM}$ \ge 10$) in the placenta of all organisms
  studied with housekeeping
  genes \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
  removed.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

plot.label.size <- 4
pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
                 offset.ratio=0.8,
                 fontsize=plot.label.size
                 )

>>>>

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental
mammals”. The term “placental mammals” is misleading, as extant
yinothereians (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, in which the maternal and fetal blood supply is separated by
one tissue layer, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
\DLA{To discuss maternal/fetal split with derek}

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Nannospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

                             "/Gene/Summary?g=",name_or_id,"}{\\textit{",
                             xtable.sanitize.text(name_or_id),"}}")]

@@

         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{\\textit{",xtable.sanitize.text(`Approved Symbol`),"}}")]

@@

    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","SEM"))
    if (temp[,all(is.na(`SEM`))]) {
        temp[,`SEM`:=NULL]

@@

placenta.lineage.specific.table <- function(analysis,output=TRUE) {
    temp.table <- 
        head(placenta.lineage.specific.analysis[analysis.name==analysis &
                                                p.value <= 0.05][order(p.value),
                                                                 list(human_name,p.value,fdr,fc),],
             n=5)
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[order(`p (t-test)`),]
        display.table <-
            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
                                                  "}")]
        print(xtable(display.table,
                     display=c("s","s","s","e","g","f"),
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

    }
    return(invisible(temp.table[order(`p (t-test)`),]))
}
output.top.genes.table <- function(data,output=TRUE) {
    if (output) {
        table.display <- c("s","s","s","d")
        table.align <- c("c","c","p{3in}","r")
        if (any(colnames(data) %in% "SEM"))  {
            table.display <- c(table.display,"d")
            table.align <- c(table.align,"r")
        }
        if (any(colnames(data) %in% "Human Symbol"))  {
            table.align <- c("c",table.align)
            table.display <- c("s",table.display)
        }
        print(xtable(data,
                     display=table.display,
                     align=table.align
                     ),
              booktabs=TRUE,
              floating=FALSE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)
    }
}

@@

\begin{figure}
<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=16,fig.height=16,warning=FALSE,message=FALSE>>=
plot.label.size <- 3
axis.text.size <- 3.5

pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
                 axis.text.size=axis.text.size,
                 fontsize=plot.label.size,
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
                 fontsize=4,
                 width=3,
                 axis.text.size=axis.text.size,
                 offset.ratio=1
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
                 fontsize=4,
                 width=3,
                 axis.text.size=axis.text.size,
                 offset.ratio=1.2
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSSSCG00000006792",
                           "ENSBTAG00000046085"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
                 axis.text.size=axis.text.size,
                 fontsize=plot.label.size,
                 offset.ratio=0.6
                 )

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (\textit{GH1} 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\textit{GH1} 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
\textit{GH2}
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
and \textit{GH2}
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
the hormone adrenomedullin (\textit{ADM}).

@@

For the term placenta of \smonkey, assembled contigs from RNA-seq were
aligned using against \human and the best alignment was chosen as the
annotation for each assembled contig. In the term placenta of
\smonkey, the top 50 non-housekeeping genes
(\cref{tab:top50_spider_monkey_placenta}) include a possible regulator
of androgen receptor signaling
\cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
(\Sexpr{gene.expression("ateles_fusciceps","PAGE4")}), %
the pregnancy specific glycoprotein \textit{PSG8}
(\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
only one member of the chorionic somatotropins, \textit{GH1} %
(\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
\cref{tab:top50_spider_monkey_placenta}), and multiple components of
annexin complexes
(\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("ateles_fusciceps","S100P",complete=TRUE)}, %
\Sexpr{gene.expression("ateles_fusciceps","S100A11",complete=TRUE)}) %
which may play a role in angiogenesis, and certainly regulate
cell-cell interaction in the developing placenta
\cite{Liu.ea2015:AnnexinA2complexeswith}.
\Sexpr{gene.expression("ateles_fusciceps","AL365202.1",complete=TRUE)}
is also highly expressed, has no known function, and is duplicated
multiple times in \textit{Macaca mulatta} (Ensembl gene tree
\href{http://useast.ensembl.org/Multi/GeneTree/Image?gt=ENSGT00580000081348}{ENSGT00580000081348}).
All genes with expression greater than 50 \ac{FPKM} are listed in
Supplementary \cref{tab:all_ateles_fusciceps_placenta}.

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \carmeli are shown in
\cref{tab:top50_s_carmeli_placenta} with a list of all highly
expressed genes in Supplemental
\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
ware highly expressed
(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Hba-a1",complete=TRUE)}, and
\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
well as the relaxin
\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
multiple cathespins
(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), and many prolactin-related genes
(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
expansion seen in \mouse.

@@

We examined the expression of five Laurasiatherians: \cow, \sheep,
\pig, \horse, and \dog. We were unable to detect any significant
differences after correcting for multiple testing between
Laurasiatherians and all other analyzed species
(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
Laurasiatherians and Euarchontoglires
(Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).

@@

\subsection{Atlantogenata (Afrotheria + Xenartha)}
\subsubsection{\elephant Placenta Transcriptome}

@@

\begin{figure}
<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
      + geom_raster(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low = grad.def[["low"]],
                            high = grad.def[["high"]],
                            space = grad.def[["space"]],
                            na.value = grad.def[["na.value"]],
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + ylab("Human Ortholog Name")
      + xlab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(),
              axis.text.y = element_text(face="italic"),
              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
      )
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes expressed (\ac{FPKM}$ \ge 10$) in the placenta of all organisms
  studied with housekeeping
  genes \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
  removed.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

plot.label.size <- 4.5
pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
                 offset.ratio=0.7,
                 fontsize=plot.label.size,
                 axis.text.size=3.5
                 )
